---
title: "Getting Your Feet Wet with R"
author: "Paul M. Magwene"
date: "August 31, 2016"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
    highlight: default 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, eval = FALSE)
```

# <font color='red'> IMPORTANT-- Read this now! </font>

In this and many future documents to be posted on the course website you'll encounter blocks of R code.  Your natural intuition will be to cut and paste commands or code blocks into the R interpretter to save yourself the typing.  **DO NOT DO THIS!!** 

Part of learning any new skill is making mistakes, figuring out where you went wrong, and correcting those mistakes.  If you cut and paste the examples into the R interpretter the code will run, but you won't learn anything.  Instead, you should type in all examples by hand, thinking about what the code means as you do so.  You'll inevitably make erorrs, generate syntax errors, etc.  In the process of fixing those errors, you'll learn more about how R works, and how to avoid such errors, or correct bugs in your own code in the future.


# What is R?

R is a statistical computing environment and programming language.  It is free, open source, and has a large and active community of developers and users.  There are many different R packages (libraries) available for conducting out a wide variety of different analyses, for everything from genome sequence data to geospatial information.

# What is RStudio?

RStudio (http://www.rstudio.com/) is an open source integrated development environment (IDE) that provides a nicer graphical interface to R than does the default R GUI.

In class we'll review the key parts of the RStudio interface.  Make sure you know where to find the console, the file explorer, the help browser, and the history and environment tabs. You should also learn how to change the look and feel of the RStudio interface.

# Accessing the help system

R comes with fairly extensive documentation and a simple help system. You can access HTML versions of the R documentation under the Help tab in Rstudio. The HTML documentation also includes information on any packages you've installed.Take a few minutes to browse through the R HTML documentation.

The help system can be invoked from the R console  using the
`help` function or the `?` operator.

```{r}
help(length)
?length
?log
```

What if you don't know the name of the function you want? You can use the `help.search()` function.

```{r}
help.search("log")
```

In this case `help.search("log")` returns all the functions with
the string `log` in them. For more on `help.search` type
`?help.search`. Other useful help related functions include
`apropos()` and `example()`.


# Comments

When working in the R console, or writing R code, the pound symbol (`#`) indicates the start of a comment.  Anything after the `#`, up to the end of the current line, is ignored by the R interpretter.

```{r, eval = FALSE}
# This line will be ignored
5 + 4 # the first part of this line, up to the #, will be evaluated
```


# Using R as a Calculator

The simplest way to use R is as a fancy calculator.

```{r, echo=TRUE, eval=FALSE}
10 + 2 # addition
10 - 2 # subtraction
10 * 2 # multiplication
10 / 2 # division
10 ^ 2 # exponentiation
10 ** 2 # alternate exponentiation
pi * 2.5^2 # R knows about some constants such as Pi
10 %% 3 # modulus operator -- gives remainder after division
10 %/% 3 # integer division
```

Be aware that certain operators have precedence over others. For example multiplication and division have higher precedence than addition and subtraction. Use parentheses to disambiguate potentially confusing statements.

```{r, eval = FALSE}
(10 + 2)/4-5   # was the output what you expected?
(10 + 2)/(4-5) # compare the answer to the above
```

Division by zero produces an object that represents infinite numbers

```{r, eval=FALSE}
0/1
1/0
```


## Basic mathematical functions

Many commonly used mathematical functions are built into R.  Here are some examples

```{r, eval=FALSE}
abs(-3)   # absolute value
abs(3)
cos(pi/3) # cosine
sin(pi/3) # sine
log(10)   # natural logarithm
log10(10) # log base 10
log2(10) # log base 2
exp(1)    # exponential function
sqrt(10)  # square root
sqrt(pi^2) 
10 ^ 0.5  # same as square root
sqrt(-1)
sqrt(-1+0i)
```


What happened when you tried to calculate `sqrt(-1)`?, -1 is
treated as a real number and since square roots are undefined for the
negative reals, R produced a warning message and returned a special
value called `NaN` (Not a Number). Note that square roots of
negative complex numbers are well defined so `sqrt(-1+0i)`
works fine.
.

## Comparison Operators

You've already been introduced to the most commonly used arithmetic
operators. Also useful are the comparison operators. Comparisons return Boolean values (`TRUE` or `FALSE`).

```{r, eval = FALSE}
10 < 9  # less than
10 > 9  # greater than
10 <= (5 * 2) # less than or equal to
10 >= pi # greater than or equal to
10 == 10 # equals
10 != 10 # does not equal
10 == (sqrt(10)^2) # Surprised by the result? See below.
4 == (sqrt(4)^2) # Even more confused?
```

Be careful to distinguish between `==`(tests equality) and `=` (one of the assignment operator).

How about the last two statement comparing two values to the square of
their square roots? Mathematically we know that both
$(\sqrt{10})^2 = 10$ and $(\sqrt{4})^2 = 4$ are true statements. Why
does R tell us the first statement is false? What we're running into
here are the limits of computer precision. A computer can't represent
$\sqrt 10$ exactly, whereas $\sqrt 4$ can be exactly represented.
Precision in numerical computing is a complex subject and beyond the
scope of this course. Later in the course we'll discuss some ways of
implementing sanity checks to avoid situations like that illustrated
above.

# Assignment operators in R

In all of the examples up until this point, we carried out some calculation and the result was immediately printed to the console.  That's convenient when we're using R as a calculator, but in any sort of more sophisticated numerical or statistical analysis we usually want to have a way of holding on to the value of our calculation.  To do so we can give our results names, like `x` or `y` or `my.long.and.silly.name`.  These names are referred to as "variables" and the process of giving names to values is called "variable assignment" or "assignment" for short.  In R, the primary assignment operators is `<-` (think of a left arrow):

```{r, eval=FALSE}
x <- 10 # assign the value 10 to x
y <- 5  # assign the value 5 to y
z <- x + y  # assign the sum of x and y to z
```

Once you've assigned a value to a variable you can carry out calculations with the variable rather than the original value.

```{r,eval=FALSE}
(x + y)^2
sqrt(x^2 + y^2)
```

Assignment also works with a right arrow, `->` but it is relatively uncommon to see this used:
```{r,eval=FALSE}
10.0 -> w
w
```

## Other assignment operators

The equal symbols (`=`) is also an assignment operator. However, there are some subtle differences between `<-` and `=` that can lead to confusion. In this class we'll use `<-` as the primary assignment operator, and reserve the equal symbol  to assign values to arguments in function calls (more on this when we discuss writing functions). 


# Working with Vectors in R

Vectors are the core data structure in R. Vectors store an ordered list of items all of the same type. Learning to compute effectively with vectors and one of the keys to efficient R programming.  Vectors in R always have a length (accessed with the `length()` function) and a type (accessed with the `typeof()` function).  

The simplest way to create a vector at the interactive prompt is to use the `c()` function, which is short hand for "combine" or "concatenate".

```{r, eval=FALSE}
x <- c(2,4,6,8)  # a vector of numbers
length(x)
typeof(x)
y <- c('joe','bob','fred') # vector of characters
length(y)
typeof(y)
z <- c() # empty vector
length(z)
typeof(z)
```

You can also use `c()` to concatenate two or more vectors together.

```{r, eval=FALSE}
v <- c(1,3,5,7)
w <- c(-1, -2, -3)
x <- c(2,4,6,8)
vwx <- c(v,w,x)
vwx
```



## Vector Arithmetic and Comparison

The basic R arithmetic operations work on vectors as well as on
single numbers (in fact single numbers *are* vectors behind the scenes).

```{r, eval=FALSE}
x <- c(2, 4, 6, 8, 10)
x * 2
x * pi
y <- c(0, 1, 3, 5, 9)
x + y
x * y
x/y
z <- c(1, 4, 7, 11)
x + z
```

When vectors are not of the same length R "recycles" the elements of the
shorter vector to make the lengths conform. In the example above
`z` was treated as if it was the vector `(1, 4, 7, 11, 1)`.


The comparison operators also work on vectors as shown below.
Comparisons involving vectors return vectors of booleans.

```{r, eval=FALSE}
x > 5
x != 4
```

If you try and apply arithmetic operations to non-numeric vectors, R will warn you of the error of your ways:

```{r, eval=FALSE}
w <- c('foo', 'bar', 'baz', 'qux')
w**2
```

Note, however that the comparison operators can work with non-numeric vectors. The results you get will depend on the type of the elements in the vector.

```{r, eval=FALSE}
w <- c('foo', 'bar', 'baz', 'qux')
w == 'bar'
w < 'cat'
```

## Indexing Vectors

For a vector of length $n$, we can access the elements by the indices $1
\ldots n$. We say that R vectors (and other data structures like lists) are `one-indexed'. Many other programming languages, such as Python, C, and Java, use zero-indexing where the elements of a data structure are accessed by the indices $0 \ldots n-1$. Indexing errors are a common source of bugs. When moving back and forth between different programming languages keep the appropriate indexing straight!

Trying to access an element beyond these limits returns a special
constant called `NA` (Not Available) that indicates missing or non-existent
values.

```{r,eval=FALSE}
x <- c(2, 4, 6, 8, 10)
length(x)
x[1]
x[4]
x[6]
```

Negative indices are used to exclude particular elements. `x[-1]` returns all
elements of `x` except the first. You can get multiple elements of a vector by
indexing by another vector. In the example above `x[c(3,5)]` returns the third
and fifth element of x`.

```{r,eval=FALSE}
x[-1]
x[c(3,5)]
```

## Combining Indexing and Comparison

A very powerful feature of R is the ability to combine the comparison
operators with indexing. This facilitates data filtering and subsetting.
Some examples:

```{r}
x <- c(2, 4, 6, 8, 10)
x[x > 5]
x[x < 4 | x > 6]
```

In the first example we retrieved all the elements of `x` that are larger than
5 (read as "x where x is greater than 5"). In the second example we retrieved
those elements of `x` that were smaller than four  *or* greater than six.
The symbol `|` is the "logical or" operator. Other logical operators
include `&` ("logical and" or "intersection") and `!` (negation). Combining
indexing and comparison is a powerful concept and one you'll probably find
useful for analyzing your own data.

## Generating Regular Sequences

Creating sequences of numbers that are separated by a specified value or
that follow a particular patterns turns out to be a common task in
programming. R has some built-in operators and functions to simplify
this task.

```{r}
s <- 1:10
s
s <- 10:1
s
s <- seq(0.5,1.5,by=0.1)
s
s <- seq(0.5, 1.5, 0.33) # 'by' is the 3rd argument so you don't have to specify it
s
```

`rep()` (short for repeat) is another way to generate patterned data.

```{r}
rep(c("Male","Female"),3)
rep(c(T,T, F),2)
```


## Creating longer vectors

For vectors of more than 10 or so elements it gets tiresome and error
prone to create vectors using `c()`. For medium length vectors
the `scan()` function is very useful.

```
> test.scores <- scan()
1: 98 92 78 65 52 59 75 77 84 31 83 72 59 69 71 66
17:
Read 16 items
> test.scores
 [1] 98 92 78 65 52 59 75 77 84 31 83 72 59 69 71 66
```

When you invoke `scan()` without any arguments the function
will read in a list of values separated by white space (usually spaces
or tabs). Values are read until |scan()| encounters a blank
line or the end of file (EOF) signal (platform dependent). We'll see how to read in data from files in the next class.

Note that we created a variable with the name `test.scores`.
If you have previous programming experience you might be surprised that
this works. Unlike most languages, R allows you to use periods in
variable names. Descriptive variable names generally improve readability
but they can also become cumbersome (e.g.
`my.long.and.obnoxious.variable.name`). As a general rule of
thumb use short variable names when working at the interpreter and more
descriptive variable names in functions.

## Useful numerical function that operate on vectors

All of the mathematical functions that we introduced earlier work with vectors:

```{r}
x <- c(0, pi/2, pi, 3*pi/2, 2*pi)
sin(x)
cos(x)
y <- c(2, 4, 6, 8)
y^2
w <- c(-1, 2, -3, 3)
abs(w)

```

Here are some additional numerical functions that are useful for
operating on vectors.

```{r}
sum(test.scores)  # test scores defined using scan() as above
min(test.scores)
max(test.scores)
range(test.scores) # min, max returned as a vector of length 2
sorted.scores <- sort(test.scores)
sorted.scores
```

# Navigating Directories in R

When you start the R environment your "working directory" (i.e. the
directory on your computer's file system that R currently "sees")
defaults to a specific directory. On Windows this is usually the same directory that R is installed in, on OS X it is typically your home directory. Here are examples showing how you can get information about your working directory and change your working directory.

```
> getwd()
[1] "/Users/pmagwene"
> setwd("/Users")
> getwd()
[1] "/Users"
```

You can also change your working directory by using the
`Set Working Directory` item under the `Session` menu in R Studio.

To get a list of the file in your current working directory use the
`list.files()` function.

```
> list.files()
[1] "Shared" "pmagwene"
```