---
title: "Bio 204: Problem Set 04"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# <font color = "red">Due Date: 24 October 2016</font>


# Problem 1: Dice rolls in role playing games

"Dungeons and Dragons" (D&D) is a table top role-playing games designed in the 70s. D&D uses a 20-sided die to decide the outcomes of combat or feats of skill.

## Problem 1.1 (2 pts)

During a battle, what is the probability of rolling a 15 or higher under the D&D rules?

## Problem 1.2 (3 extra credit)

"Tunnels and Trolls" (T&T) is another fantasy role-playing game from the 70s. T&T uses three 6-sided dice to decide the outcomes of combat or feats of skill. What is the probability of rolling a 12 or better under the T&T rules?

# Problem 2 (5 pts)

A 2012 Pew Research survey asked 2,373 randomly sampled registered voters their political affiliation (Republican, Democrat, or Independent) and whether or not they identify as swing voters. 35% of respondents identified as Independent, 23% identified as swing voters, and 11% identified as both.

(a) Are being Independent and being a swing voter disjoint, i.e. mutually exclusive?

(b) What percent of voters are Independent but not swing voters?

(c) What percent of voters are Independent or swing voters?

(d) What percent of voters are neither Independent nor swing voters?

(e) Is the event that someone is a swing voter independent of the event that someone is a political Independent?




# Problem 3: The `births` data

The `births.txt` data set is available at the following URL:

https://raw.githubusercontent.com/Bio204-class/bio204-datasets/master/births.txt

The `births` data set contains information on 150 cases of mothers and their newborns in North Carolina. The variables in the data set are: mother's age (mAge), father's age (fAge), weeks of gestation (weeks), whether the birth was premature or full term (premature), number of OB/GYN visits (visits), mother's weight gained in pounds (weight), sex of the baby (sexBaby), and whether the mother was a smoker (smoke).



## Problem 3.1 (3 pts)

Use the `table` function to create a contingency table (see OpenIntro Stats p. 43 and p. 92) from the `births` data set, summarizing the joint counts for the categorical variables `premature` and `smoke`.

Your output should look something like this, with the correct values substituted for `n1`, `n2`, etc.

```
         nonsmoker smoker
  full term        n1     n2
  premie           n3     n4
```

```{r}
# Problem 3.1 code here
```

## Problem 3.2 (3 pts)

The `addmargins` function can be used to add marginal sums to a matrix or table.  Use `addmargins` to add the marginal sums to the contingency table you created above.  Your output should look something like this (again with the appropriate sums in each cell):

```
           nonsmoker   smoker   Sum
  full term        n1      n2   (n1+n2)
  premie           n3      n4   (n3+n4)  
  Sum         (n1+n3) (n2+n4)   (n1+n2+n3+n4)
```

```{r}
# Problem 3.2 code here
```

## Problem 3.3 (6 pts)

From the table you created in 1.2, calculate the following probabilities:

(a) $P(\mathrm{smoker})$

(b) $P(\mathrm{premie})$

(c) $P(\mathrm{smoker} \cap \mathrm{premie})$

(d) $P(\mathrm{nonsmoker} \cap \mathrm{premie})$

(e) $P(\mathrm{premie}|\mathrm{nonsmoker})$

(f) $P(\mathrm{premie}|\mathrm{smoker})$


# Problem 4 (4 pts)

Lupus is a disease where antibodies that are supposed to attack foreign cells to prevent infections instead see plasma proteins as foreign bodies, leading to a high risk of blood clotting. It is believed that 2% of the population suffer from this disease. The test is 98% accurate if a person actually has the disease. The test is 74% accurate if a person does not have the disease. There is a line from the  television show "House" that is often used after a patient tests positive for lupus: “It’s never lupus.” (see https://www.youtube.com/watch?v=bueW1i9kQao). Do you think there is truth to this statement? Use appropriate probabilities to support your answer.

```{r}
# R code in support of your calculations
```
