---
title: "Problem Set 02"
author: "Paul M. Magwene"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DUE DATE: 28 September 2016


# Problem 1 (2 pts)

Pick four functions from the ggplot library that we have not discussed in class, write brief descriptions (in your own words) of what these functions do, and then provide brief code example illustrating their use (don't simply repeat the examples from the documentation!).  Use any of the data set examples we have discussed in class in your code, or consider data sets from the [R Datasets Package](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html) (the Datasets package is installed and loaded by default in a stock R setup).

### Function 1 description

```{r}
### Function 1 example
```

### Function 2 description

```{r}
### Function 2 example
```

### Function 3 description

```{r}
### Function 3 example
```

### Function 4 description

```{r}
### Function 4 example
```



# Problem 2

The file [possums.txt](https://github.com/Bio204-class/bio204-datasets/raw/master/possum.txt) is a tab-delimited file that contains morphological measurements on 104 brushtail possums (*Trichosurus vulpecula*). This data set contains information on variables such as sex, age, head length, skull width, tail length, and total body length, as well as information about the population and locality (site) at which each specimen was collected.


## Problem 2.1: Creating summary tables (3 pts)

Use the `dplyr` functions `group_by` and `summarise`, and the built-in function `as.data.frame`, to generate the table below, which gives the mean values of each of the four morphological variables (`headL`, `skullW`, `totalL`, and `tailL`) in the two populations represented in the data.

Your expected output should look like:
```
    pop mean.headL mean.skullW mean.totalL mean.tailL
1 other   92.60690    57.06552    86.78793   37.86207
2   Vic   92.59783    56.65435    87.46739   35.93478
```

```{r}
# Problem 2.1 code
```


## Problem 2.2: Boxplots (4 pts)

Using ggplot, recreate the boxplot given below that shows the tail length of possums, conditional on population and sex.

```{r}
# Problem 2.2 code
```
