---
title: "R Data Frames"
author: "Paul M. Magwene"
date: "September 5, 2016"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
    highlight: default  
    fig_width: 5
    fig_height: 3    
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, eval = FALSE, 
                      comment=NA, warning = FALSE, results="hide")
```

# Data Frames

Along with vectors and lists, data frames are one of the core data structures
when working in R.  A data frame is essentially a list which represents a data
table, where each column in the table has the same number of rows. Also unlike standard lists, the objects (columns) in a data frame must have names.

## Creating a data frame in the console

```{r, eval=TRUE}
age <- as.integer(rnorm(10, mean = 30, sd = 5))
sex <- rep(c("M","F"), 5)
wt <- rnorm(10, mean=50, sd = 8)
data <- data.frame(age = age, sex=sex, weight=wt)
data  # examine the data frame
```

## Indexing and accessing data in data frames

Once you have a data frame object, you can access the elements in the data frame by either row, column, or both. Here are some examples

```{r, eval=TRUE, results="hide"}
data$age       # get the age column
data[c("age", "weight")] # get the age and weight columns
data[1:2,]     # get the first two rows
data[1:3, 1:2] # get the first three rows and first three columns
```

## Reading files as a data frame

Most functions for reading data from files, such as `read.csv` and `read.table` typically return the data they read as data frames.  We previously examined the
`read.csv` function which parses "Comma Separated Value" format.  The 
`read.table` function does the same for tab or space delimited files.

The files `turtles.txt` (from the course GitHub website) is a tab-delimited file with a set of measurements representing linear dimensions of the carapace (upper shell) of painted turtles (*Chrysemys picta*), as reported in Jolicoeur and Mosimmann, 1960; Growth 24: 339-354.
