---
title: "Visualizing and Describing Bivariate Distributions"
author: "Paul M. Magwene"
date: "September 19, 2016"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
    highlight: default  
    fig_width: 5
    fig_height: 3.25
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = TRUE, 
                      comment=NA, warning = FALSE, results="hide",
                      message = FALSE, cach = TRUE)
```

# Introduction

Having already explored various ways to visualize the distributions of single variables, we now turn our attention to bivariate and multivariate representations.  

```{r}
library(dplyr)
library(ggplot2)
```

# Example data set: Iris

Initially we'll use just the I. setosa specimens from the Iris data set for our visualizations.

```{r}
setosa <- filter(iris, Species == "setosa")
```


# Scatter plots

A scatter plot is one of the simplest representations of a bivariate distribution. 

Scatter plots are simple to create in ggplot simply by specifying the appropriate X and Y variables in the aesthetic mapping and using `geom_point` for the geometric mapping.

```{r}
ggplot(setosa, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
```

I'm not crazy about the defaults in ggplot, so let's tweak our plot a little:

```{r}
setosa.length.vs.width <- ggplot(setosa, aes(x = Sepal.Length, y = Sepal.Width))

clean.theme <- theme_bw() +                 # removes the background
  theme(panel.grid.major = element_blank()) # remove the heavy grid lines

labels <- labs(x = "Sepal Length (cm)", y = "Sepal Width (cm)")

setosa.length.vs.width + 
  geom_point(alpha = 0.25, size = 2) + # tweak alpha and size of points
  labels + clean.theme
```

You'll recall that there was a fair amount of overlap between the observations -- you can tell this is the case because when we set the alpha transparency in the previous figure some of the plotted points look darker than other. We can substitute `geom_jitter` for `geom_point` if we wish to add jitter to the points in our scatter plot.

```{r}
setosa.length.vs.width + 
  geom_jitter(alpha = 0.25, size = 2,           # tweak shape and size
              width = 0.05, height = 0.05) +    # set degree of jitter
  labels + clean.theme
```

## Setting aspect ratios

When the X and Y variables of a scatter plot are measured on the same scale and have comparable variance it's sometimes useful to draw them plot with an aspect ratio of 1, meaning that a unit distance in the x-direction is the same as unit distance in the y-direction.  

```{r}
setosa.length.vs.width + geom_point(alpha = 0.25, size = 2) + labels + clean.theme +
  theme(aspect.ratio = 1) +       # set aspect ratio
  xlim(3.5,6.5) + ylim(2,5)       # set limits for x- and y- axes
```

# Marginal histograms and density plots with ggExtra

The ggExtra package provides some useful additional functions for working with ggplot. We'll look at a function called `ggMarginal` that provides a convenient way to plot histograms or density plots on the margin of a scatter plot.  First you'll need to install `ggExtra` using the RStudio GUI or the command `install.packags("ggExtra")`.

```{r}
library(ggExtra)

p <- setosa.length.vs.width + geom_point(alpha = 0.25, size = 2) + 
  labels + clean.theme + theme(aspect.ratio = 1) 

ggMarginal(p, type = "histogram", bins = 9)
```

ggMarginal can also generate marginal density plots and boxplot.

```{r}
ggMarginal(p, type = "density")
```

# Bivariate density plot

The density plot, which we introduced as a visualization for univariate data, can also be depicted in two-dimensions.  It 2D density plot has the same interpretation as in 1D, but not with respect to a bivariate distribution.  The contours of the 2D density plot indicate contours of estimated equal density.

```{r}
setosa.length.vs.width + geom_density2d() + labels + clean.theme
```

We can make the relationship between the 2D density plot and the scatter plot can be made clearer if we combine the two:

```{r}
setosa.length.vs.width + geom_density_2d() + geom_jitter(alpha=0.35) +
  labels + clean.theme + theme(aspect.ratio = 1)
```

# Combining Scatter Plots and Density Plots with Categorical Information

As with many of the univariate visualizations we explored, it is often useful to depict bivariate relationships as we change a categorical variable.  Here are some examples of this:

```{r}
all.length.vs.width <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width))

all.length.vs.width + 
  geom_point(aes(color = Species), size = 2, alpha = 0.6) +
  labels
```

The same thing can be accomplished with a 2D density plot.
```{r}
all.length.vs.width + 
  geom_density_2d(aes(color = Species)) +
  labels
```

# Facet plots in ggplot

As you can see, in the density plots above, when you have multiple categorical variables and there is significant overlap in the range of each sub-distribution, figures can become quite busy.  One way to deal with this is to create a separate subplot for each distribution.  ggplot provides a mechanism for doing this called a `facet_grid`.

```{r}
all.length.vs.width + geom_point(aes(color = Species)) + 
  clean.theme + theme(aspect.ratio = 1) + labels + 
  facet_grid(. ~ Species)
```

The first argument to `facet_grid` is a "formula" that specifies what is drawn with respect to the rows and columns of the grid.  The formula can be generalized as `LHS ~ RHS` (LHS = left hand side, RHS = right hand side) where the LHS gives the rows and the RHS give the columns. A dot in the formula indicates no faceting in that dimension.

Here's a combination of scatter plots and 2D density plots, embedded in a facet grid.
```{r}
all.length.vs.width + 
  geom_density_2d(aes(color = Species), alpha = 0.5) + 
  geom_point(aes(color = Species), alpha=0.5, size=1) + 
  facet_grid(. ~ Species) +
  clean.theme + labels + 
  theme(aspect.ratio = 1, legend.position = "none") + 
  labs(title = "Iris Sepal Morphology")
```
