---
title: "Snippets"
author: "Paul M. Magwene"
date: "September 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Lists in R

R lists are like vectors, but unlike a vector where all the elements are of the same type, the elements of a list can have arbitrary types (even other lists). Because vectors only hold a single type of data, they are more efficient, however lists are more flexible for representing heterogenous collections of data.

Here's examples of creating lists:
```{r}
l <- list('Bob', pi, 10, c(2,4,6,8))
m <- list(1, 2, 'three', list('a','b','c'))
```

## Indexing lists
Indexing of lists is different than indexing of vectors.  Double brackets (`x[[i]]`) return the element at index $i$, single bracket return a list containing the element at index $i$.

```{r}
l[1] # single brackets
l[[1]] # double brackets
typeof(l[1])  # returns another list
typeof(l[[1]]) # returns the element itself
```

## Getting parts of lists by name
The elements of a list can be given names, and those names objects can be accessed using the `$` operator. You can retrieve the names associated with a list using the `names()` function.

```{r}
p <- list(first.name='Alice', last.name="Qux", age=27, years.in.school=10)
p$years.in.school
p$first.name
p$last.name
names(p)
```

# Simple Input in R

The `c()` and `scan()` functions are fine for
creating small to medium vectors at the interpreter, but eventually
you'll want to start manipulating larger collections of data. There are
a variety of functions in R for retrieving data from files.

The most convenient file format to work with are tab delimited text
files. Text files have the advantage that they are human readable and
are easily shared across different platforms. If you get in the habit of
archiving data as text files you'll never find yourself in a situation
where you're unable to retrieve important data because the binary data
format has changed between versions of a program.
